---
title: "Analyzing Student Learning Outcomes"
author: "Sayali Pathak"
date: "2024-12-15"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
# Set CRAN mirror \
options(repos = c(CRAN = "https://cloud.r-project.org"))

# First, install required packages if they're not already installed
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("DBI")) install.packages("DBI")
if (!require("RMySQL")) install.packages("RMySQL")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("dplyr")) install.packages("dplyr")
if (!require("tidyr")) install.packages("tidyr")
if (!require("corrplot")) install.packages("corrplot")
if (!require("lme4")) install.packages("lme4")
if (!require("knitr")) install.packages("knitr")
if (!require("scales")) install.packages("scales")

# load the packages
library(tidyverse)
library(DBI)
library(RMySQL)
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
library(lme4)
library(knitr)
library(scales)

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)


# Setting global options for better visualization
theme_set(theme_minimal())
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Data Collection and Preparation
```{r data_prep}

con <- dbConnect(MySQL(),
                dbname = "StudentOutcomes",
                host = "localhost",
                user = "root",
                password = "3004")


query <- "
SELECT 
    sp.*,
    CASE 
        WHEN weekly_self_study_hours >= 25 THEN 'High'
        WHEN weekly_self_study_hours >= 15 THEN 'Medium'
        ELSE 'Low'
    END as study_intensity,
    (math_score + physics_score + chemistry_score) / 3 as stem_average,
    (history_score + english_score + geography_score) / 3 as humanities_average
FROM StudentPerformance sp
WHERE absence_days <= 10"

students_raw <- dbGetQuery(con, query)
dbDisconnect(con)

# Tidyverse data cleaning
students <- students_raw %>%
  tidyr::drop_na() %>%
  dplyr::mutate(
    across(ends_with("_score"), ~ scale(.) %>% as.vector),
    study_intensity = factor(study_intensity, levels = c("Low", "Medium", "High")),
    engagement_level = case_when(
      extracurricular_activities & !part_time_job ~ "Full Academic",
      extracurricular_activities & part_time_job ~ "Balanced",
      !extracurricular_activities & part_time_job ~ "Work Focused",
      TRUE ~ "Minimal Engagement"
    )
  )
```

## Learning Analytics Insights

### 1. Study Pattern Impact Analysis
```{r study_patterns}
# Advanced ggplot2 visualization
ggplot(students, aes(x = weekly_self_study_hours, y = stem_average)) +
  geom_point(aes(color = engagement_level), alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~gender) +
  labs(
    title = "Impact of Study Hours on STEM Performance",
    subtitle = "Analyzed by Gender and Engagement Level",
    x = "Weekly Self-Study Hours",
    y = "Standardized STEM Score Average",
    color = "Engagement Type"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "italic")
  )
```

### 2. Hierarchical Learning Pattern Analysis
```{r hierarchical}

hier_model <- lmer(stem_average ~ 
                    weekly_self_study_hours + 
                    (1 + weekly_self_study_hours | career_aspiration) +
                    (1 | engagement_level),
                  data = students)

# Create accessible summary for varying data literacy levels
model_summary <- summary(hier_model)

# Visualization of random effects
ranef(hier_model)$career_aspiration %>%
  as.data.frame() %>%
  rownames_to_column("Career") %>%
  ggplot(aes(x = reorder(Career, weekly_self_study_hours), y = weekly_self_study_hours)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Study Hours Effect by Career Aspiration",
    subtitle = "From Hierarchical Model Analysis",
    x = "Career Path",
    y = "Effect Size"
  )
```

### 3. Educational Equity Analysis
```{r equity}
# Creating an equity index
students %>%
  group_by(gender, engagement_level) %>%
  summarise(
    avg_stem = mean(stem_average),
    avg_humanities = mean(humanities_average),
    opportunity_gap = abs(avg_stem - avg_humanities),
    .groups = 'drop'
  ) %>%
  ggplot(aes(x = engagement_level, y = opportunity_gap, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Educational Opportunity Gap Analysis",
    subtitle = "By Engagement Level and Gender",
    x = "Engagement Level",
    y = "Performance Gap (Standardized)"
  ) +
  scale_fill_brewer(palette = "Set2")
```

## Recommendations for Educational Impact


1. **Differentiated Support**: Structure support systems based on engagement patterns
2. **Equity Interventions**: Address identified opportunity gaps
3. **Resource Optimization**: Align study resources with career aspirations
4. **Engagement Strategies**: Develop programs that balance academic and extracurricular activities

## Methodology Notes

This analysis employs:
- Hierarchical linear modeling for nested educational data
- Robust visualization techniques for diverse stakeholders
- Educational equity metrics
- Evidence-based recommendation framework

```{r save_report}
# Save visualizations for stakeholder presentation
ggsave("study_impact.png", width = 10, height = 6)
```